
```{r}
#| include: false
#load and preview dataset-asah

library(pROC)
library(ggplot2)
library(mlbench)

data(aSAH)
head(aSAH)
str(aSAH)

#load-heart
heart <- read.csv(
  "https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data",
  header = FALSE,
  na.strings = "?"
)
colnames(heart) <- c("age", "sex", "cp", "trestbps", "chol",
                     "fbs", "restecg", "thalach", "exang",
                     "oldpeak", "slope", "ca", "thal", "target")
heart$target <- ifelse(heart$target > 0, 1, 0)

roc_asah <- roc(response = aSAH$outcome,
                predictor = aSAH$s100b,
                direction = "<")
```


# Compute and Interpret AUC

The Area Under the Curve (AUC) summarizes the ROC curve in a single number.

- **AUC = 0.5**: the test performs no better than random chance
- **AUC = 1.0**: the test perfectly separates diseased from non-diseased individuals
- **AUC = 0.7–0.8**: generally considered acceptable discrimination
- **AUC > 0.8**: generally considered excellent discrimination

These are rough benchmarks — what counts as "good enough" depends on your clinical or scientific context.

```{r}
#| label: auc-asah
auc(roc_asah)
ci.auc(roc_asah)
```

Repeat for the other datasets:

```{r}
#| label: roc-pima
data("PimaIndiansDiabetes")

roc_pima <- roc(response = PimaIndiansDiabetes$diabetes,
                predictor = PimaIndiansDiabetes$glucose,
                direction = "<")
auc(roc_pima)
ci.auc(roc_pima)
```

```{r}
#| label: roc-heart
heart_clean <- na.omit(heart[, c("thalach", "target")])
roc_heart <- roc(response = heart_clean$target,
                 predictor = heart_clean$thalach,
                 direction = ">")  # higher heart rate -> less likely disease
auc(roc_heart)
ci.auc(roc_heart)
```

