
```{r}
#| include: false
#load and preview dataset-asah

library(pROC)
library(ggplot2)
library(mlbench)

data(aSAH)
head(aSAH)
str(aSAH)

#load-heart
heart <- read.csv(
  "https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data",
  header = FALSE,
  na.strings = "?"
)
colnames(heart) <- c("age", "sex", "cp", "trestbps", "chol",
                     "fbs", "restecg", "thalach", "exang",
                     "oldpeak", "slope", "ca", "thal", "target")
heart$target <- ifelse(heart$target > 0, 1, 0)
```



# Generate the ROC Curve

Use `pROC::roc()` to fit the ROC curve. The two key arguments are:

- `response`: the binary outcome variable
- `predictor`: the continuous test result

```{r}
#| label: roc-asah
roc_asah <- roc(response = aSAH$outcome, predictor = aSAH$s100b)
roc_asah
```

## Understanding the Direction Warning

`pROC` will print a message about the `direction` it assumed (e.g., `direction: controls < cases`). This tells you whether higher test values are assumed to indicate disease. **Do not ignore this.** If the assumed direction is wrong for your test, your curve will be mirrored and your AUC will be below 0.5. You can set the direction explicitly:

```{r}
#| label: roc-direction
# Higher s100b -> worse outcome: controls < cases (default is correct here)
roc_asah <- roc(response = aSAH$outcome,
                predictor = aSAH$s100b,
                direction = "<")
```
